---
title: "Module API and index lifecycle"
---

If you want to add items to an index over time, delete them, or save the index for later, you can
use the Rcpp Module classes directly. The usual [`hnsw_build()`](../reference/hnsw_build.html) and
[`hnsw_search()`](../reference/hnsw_search.html) functions use these same classes, so an index
you've already built with `hnsw_build()` has the methods described here too.

# Create an index

| Class | Distance |
| --- | --- |
| `HnswEuclidean` | Euclidean |
| `HnswL2` | Squared Euclidean |
| `HnswCosine` | One minus cosine similarity |
| `HnswIp` | One minus inner product |

All four classes have the same constructors and methods. We'll use `HnswL2` here and leave room for
100 items to begin with. The constructor takes the dimension, capacity, `M`, construction `ef`, and
an optional random seed (default `100`), in that order. The order matters: Module arguments are
matched by position, even if you give them names.

```{r}
library(RcppHNSW)

items <- as.matrix(iris[, -5])
dimension <- ncol(items)
capacity <- 100
M <- 16
ef_construction <- 200
ann <- new(HnswL2, dimension, capacity, M, ef_construction, 100)
ann$addItems(items[1:100, ])
```

For help choosing `M`, construction `ef`, and the seed, see the
[parameter guide](parameters-metrics-data-layout.html).

# Add and retrieve items

| Method | Input or result |
| --- | --- |
| `addItem(v)` | Add one vector. |
| `addItems(m)` | Add each row of a matrix. |
| `addItemsCol(m)` | Add each column of a matrix. |
| `getItems(ids)` | Return one row per requested label, in the order supplied. |

Items get labels in the order you add them, starting at one. So `ann$getItems(c(1, 10))` retrieves
the first and tenth items. If you're using a cosine index, the returned vectors have been
normalized to unit length.

Coordinates are stored in single precision; see
[numeric limits](../reference/RcppHnsw-package.html#numeric-limits)
for the accepted range.

# Search

Set search `ef` with `ann$setEf(ef)`. As with `hnsw_search()`, increasing it usually finds more of
the true nearest neighbors, but takes longer. `k` must be positive and no greater than the number
of undeleted items. If a search cannot find `k` neighbors, try increasing `ef`; a poorly connected
index may need rebuilding with a larger `M`.

| Method | Query layout | Result |
| --- | --- | --- |
| `getNNs(v, k)` | One vector | Vector of labels |
| `getAllNNs(m, k)` | One item per row | `n × k` matrix of labels |
| `getAllNNsCol(m, k)` | One item per column | `k × n` matrix of labels |

The corresponding `getNNsList()`, `getAllNNsList()`, and `getAllNNsListCol()` methods take a third
argument, `include_distances`. They return labels as `item` and, when that argument is `TRUE`,
distances as `distance`, with the same shapes as above.

For example, here's how to get both labels and distances for the first five items:

```{r}
ann$setEf(50)
result <- ann$getAllNNsList(items[1:5, ], 4, TRUE)
result$item
result$distance
```

Batch insertion and search use the settings from `setNumThreads(n_threads)` and
`setGrainSize(grain_size)`. See
[threads and reproducibility](parameters-metrics-data-layout.html#use-multiple-threads).

# Resize and delete

We've filled the 100 spaces we asked for, but there are another 50 items in `iris`. To add those,
we first need to make room with `resizeIndex(capacity)`. The new capacity must be at least one and
at least `size()`, which includes deleted items.

```{r}
ann$resizeIndex(nrow(items))
ann$addItems(items[101:150, ])
ann$size()
```

`markDeleted(label)` hides an item from search and `getItems()`. The "mark" part of the name is
doing some work here: the item still occupies space in the index. Deleting it doesn't free memory
or capacity, and `size()` still counts it:

```{r}
ann$markDeleted(1)
ann$size()
```

If an insertion fails after it starts modifying the index, or a native resize fails, discard the
index and rebuild or reload it. Invalid inputs rejected before modification leave the index usable.

# Save and load an index

Once you have an index you're happy with, `ann$save(filename)` saves the graph, its items, capacity,
and deletion state. Let's save ours and load it into a new object:

```{r}
path <- tempfile(fileext = ".hnsw")
ann$save(path)
loaded <- new(HnswL2, dimension, path)
loaded$setEf(50)
loaded$size()
unlink(path)
```

Use the original dimension and distance class when loading. `HnswL2` and `HnswEuclidean` can also
load each other's checkpoints. Search `ef` isn't saved: loading resets it to 10, which is why we
set it again above.

To change capacity while loading, use `new(HnswL2, dimension, path, capacity)`. Omitting capacity
restores the saved value. An explicit positive capacity takes effect when it is at least the
stored item count; smaller values fall back to the saved capacity.

These files use hnswlib's raw format, and compatibility depends on the hnswlib version and platform.
Keep the original data too, in case you need to rebuild the index elsewhere.
